Future challenges for Research Software Engineering

### 2025-08-14 correspondence/rebuttal-1.md

> The paper overall is very generic and more related to Software Engineering in general than specific to Research Software. Similar papers on Software Engineering issues already exist and provide more details on the actual challenges, expected development and which technologies and approaches may prove relevant. In most cases, this paper is leaving out concrete issues and already existing developments (see e.g. Intellisoft).

### Technical soundness

> The paper is a bit too generic for an in-depth assessment of its technical soundness. Specifically, many of the "core issues" are mentioned but never listed (with exception of "core software skills" - which ever these may be). As noted under significance, some existing developments in Software Engineering have not been properly respected (AI tooling, standardisation etc.) - which all contributed to the simplicity that allows easy generation of software. RSE has specific additional challenges related to soundness, replicability, reproducibility, interpretability etc. which are not addressed in the paper.

Thank you for this observation.
We were unaware of the SE community work on explainable AI for SE (XAI4SE) and AI for RSE (AI4RSE).
We have extended the section on AI tooling and its challenges using the SE literature on the subject.
We now mention the issue of reproducibility and draw a parallel with the HPC literature on the subject.

@@

> The paper chooses a very interesting approach to present the key challenges / aspects, with a short description of the aspect leading as a mid- to long-term goal and a "futuristic story" as an exemplary scenario. Most of these aspects relate however to communities recognizing the relevance of RSE much more than what the relevance is, and the futuristic scenarios sometimes do not seem to fit the topic of the goal, such as in the context of Ethics / Social Consideration; few of the scenarios provide additional information that would further the understanding of the challenges or goals.

### Suggestions

> I like the idea and the elaboration of the paper, but fail to see relevant contributions in the individual sections, in particular comes to differentiation from Software Engineering. With that respect, related work in Software Engineering already addresses a lot of aspects mentioned in the paper and should be referenced and acknowledged appropriately as it shapes expectations and future work. I'd love to see a better elaboration of the paper, as again it is a relevant and interesting approach.

This is a fair point.
We have familiarized ourselves with the SE literature on this subject and have drawn
parallels between projections made in the SE space and made in the RSE space.
It was particularly interesting to see the overlap between our manuscript
and Katz's 2023 paper ``Research Software Engineering in 2030''.

Many participants of the workshop are scientists by training, and therefore have slightly
different perspectives on the future of RSEng due to the environment they work in.
They also use a different vocabulary compared to software engineers.
The addition of SE literature is definitively an improvement for the manuscript.

### 2025-08-14 future_challenges_for_RSEng.tex
More SE literature

Research Software Engineering is still a developing field,
with increasing integration into various aspects of research.
In Germany, a formal call for creating a sustainable environment for research software
is only a couple of years behind~\cite{Anzt2021},
while the foundational competencies of a research software engineer have only recently been defined in detail~\cite{Goth2024}. Central RSE units are in the process of being imagined~\cite{Kempf2025-draft},
while an RSE Master's is in the early stages of design.

In this opinion paper, we try to imagine the state of our field
and community in the mid-term future, focusing on future challenges.
While future projections are common in industry~\cite{Gartner2025},
so far there is little projection for the RSEng space.
We present here the results of a workshop discussion (notes content is archived here \cite{goth2025_WSPad}) at the
deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}?
Challenges and Risks for {RSEng}'': \url{https://events.hifis.net/event/1741/contributions/14026/}},
without claiming to represent the entire deRSE community.
The material presented here is based on the output of the workshop participants, but extended with the authors' ideas.
Since deRSE25 was co-located with SE25, the annual conference for the Software Engineering community in Germany, the discussion was also open to software engineers,
but no list or profile of the participants is available.
The workshop hosts felt, that the majority of participants were from the RSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in creating a future worth living
and to advance the conversation about how Research Software Engineering should, or could, develop.

It is already difficult to structure the spectrum of RSE tasks and responsibilities
today given the spectrum of research software~\cite{Hasselbring2024}.
It is even more challenging to structure the aspects that will become relevant for the RSE community in the future.
In order to structure the discussion, we decided to consider
the evolution of the term RSE and its definition,
the growth and development of the RSE community,
the impact of automation and generative artificial intelligence in RSE work,
the impact of complexity and usability in research software,
ethical and social considerations,
the expected transformations in academic institutions,
and the potential impact of alternative career paths for RSEs
and new groups of RTPs (\emph{Research Technical Professional}s).
This work supplements other opinion pieces developed in parallel
in the SE community~\cite{ChueHong2025,Sochat2024Infra,Bencomo2024AEBoK,OliveiraJr2024,Druskat2025},
which where not known to the workshop participants; remarkably,
many of the projections made by the participants overlap with predictions made
by the SE community in ``Research Software Engineering in 2030''~\cite{Katz2023}.
We conclude the paper with a summary and with our wishes and recommendations.

@@

While the RSE core competencies~\cite{Goth2024} might remain unchanged,
there might be a shift in the responsibilities of RSEs,
in the form of a partial reallocation of the training budget
to the other core competencies.
Regarding the definition of RSEs, we will still have a spectrum of competencies,
from full professional RSEs to domain scientists who develop research software.

>>>>

Research Software Engineering is still a developing field,
with increasing integration into various aspects of research.
In Germany, a formal call for creating a sustainable environment for research software
is only a couple of years behind~\cite{Anzt2021},
while the foundational competencies of a research software engineer have only recently been defined in detail~\cite{Goth2024}.
Central RSE units are in the process of being imagined~\cite{Kempf2025-draft},
while an RSE Master's is in the early stages of design~\cite{Dehne2025-draft}.

In this opinion paper, we try to imagine the state of our field
and community in the mid-term future, focusing on future challenges.
While future projections are common in software engineering
(SE)~\cite{Katz2023,Khan2021,Hu2023,Bosch2016a,Boehm2011},
so far there is little projection for the RSEng space.
We present here the results of a workshop discussion (notes content is archived here \cite{goth2025_WSPad}) at the
deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}?
Challenges and Risks for {RSEng}'': \url{https://events.hifis.net/event/1741/contributions/14026/}},
without claiming to represent the entire deRSE community.
The material presented here is based on the output of the workshop participants, but extended with the authors' ideas.
Since deRSE25 was co-located with SE25, the annual conference for the Software Engineering community in Germany, the discussion was also open to software engineers,
but no list or profile of the participants is available.
The workshop hosts felt, that the majority of participants were from the RSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in creating a future worth living
and to advance the conversation about how Research Software Engineering should, or could, develop.

It is already difficult to structure the spectrum of RSE tasks and responsibilities
today given the spectrum of research software~\cite{Hasselbring2024}.
It is even more challenging to structure the aspects that will become relevant for the RSE community in the future.
In order to structure the discussion, we decided to consider
the evolution of the term RSE and its definition,
the growth and development of the RSE community,
the impact of automation and generative artificial intelligence in RSE work,
the impact of complexity and usability in research software,
ethical and social considerations,
the expected transformations in academic institutions,
and the potential impact of alternative career paths for RSEs
and new groups of RTPs (\emph{Research Technical Professional}s).
This work supplements other opinion pieces developed in parallel
in the SE community~\cite{ChueHong2025,Sochat2024Infra,Bencomo2024AEBoK,OliveiraJr2024,Druskat2025,Carleton2022},
which where not known to the workshop participants; remarkably,
many of the projections made by the participants overlap with predictions made
by the SE community in ``Research Software Engineering in 2030''~\cite{Katz2023}.
We conclude the paper with a summary and with our wishes and recommendations.

@@

While the RSE core competencies~\cite{Goth2024} might remain unchanged,
there might be a shift in the responsibilities of RSEs,
in the form of a partial reallocation of the training budget
to the other core competencies.
Regarding the definition of RSEs, we will still have a spectrum of competencies,
from full professional RSEs to domain scientists who develop research software.
We note a similar discussion took place regarding how SE roles had to adapt to
rapid changes brought by automation, cloud services and new project management
methodologies~\cite{Meade2019}.
